The human visual system preserves the hierarchy of 2-dimensional pattern regularity

### 2020-07-30 Writeup/manuscript.tex
updated results

% If your first paragraph (i.e. with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation. If this is the case, add \parshape=0 to the end of the list environment.
\dropcap{S}ymmetries are present at many scales in images of natural scenes, due to a complex interplay of physical forces that govern pattern formation in nature. The importance of symmetry for visual perception has been known at least since the gestalt movement of the early 20th century. Since then, symmetry has been shown to contribute to the perception of shapes \cite{RN1311,RN1682}, scenes \cite{RN1824} and surface properties \cite{RN1166}, as well as the social process of mate selection \cite{RN1337}. Most of this work has focused on mirror symmetry or \textit{reflection}, with much less attention being paid to the other fundamental symmetries: \textit{rotation}, \textit{translation} and \textit{glide reflection}. In the two spatial dimensions relevant for images, these four fundamental symmetries can be combined in 17 distinct ways, the “wallpaper” groups \cite{RN1562,RN1563,RN1425}. Previous work has focused on four of the wallpaper groups, and used functional MRI to show that rotation symmetries within wallpapers are represented parametrically in several areas in occipital cortex, beginning with visual area V3 \cite{RN1725}. This effect is also robust in electroencephalography (EEG), whether measured using Steady-State Visual Evoked Potentials (SSVEPs)\cite{RN1725} or event-related paradigms \cite{RN1959}. Here we extend on this work by collecting SSVEPs and psychophysical data from human participants viewing the complete set of wallpaper groups. We measure responses in visual cortex to 16 out of the 17 wallpaper groups, with the 17th serving as a control stimulus, with the goal of providing a more complete picture of how wallpaper groups are represented in the human visual system. The 17 wallpaper groups obey a hierarchy of complexity, determined by mathematical group theory, where simpler groups are sub-groups of more complex ones \cite{RN1711}. The two datasets presented here puts on in the position of being able to assess the extent to which both behavior and brain responses follow that hierachy. The results show that activity in human visual cortex is remarkably consistent with the hierarchical relationships between the wallpaper groups, with SSVEP amplitudes and psychophysical thresholds preserving these relationships at a level that is far beyond chance. Visual cortex thus appears to encode all of the fundamental symmetries using a representational structure that closely approximates the subgroup relationships from group theory. Given that most participants had no knowledge of group theory, the ordered structure of visual responses to wallpaper groups is likely learned implicitly from regularities in the visual environment.

@@

The visual stimuli for our experiment were multiple exemplar images belonging to each of the 17 wallpaper groups, generated from random-noise textures, as described in detail elsewhere \cite{RN1725}. To isolate brain activity specific to the symmetry structure in the images from activity associated with modulation of low-level features, we used a steady-state design, in which exemplar images belonging to 16 of the 17 wallpaper groups alternated with phase-scrambled images of the same group. Because all wallpapers are periodic images due to their lattice tiling structure, the phase-scrambled images are also a wallpaper group (P1). P1 contains no symmetries other than translation, while all other groups contain translation in combination with one or more of the other three fundamental symmetries (reflection, rotation, glide reflection) \cite{RN1425}. 
Exemplars from each of the 16 groups alternated at 0.83 Hz with their corresponding set of P1 exemplars, that were matched in terms of their Fourier power spectrum. Because the P1 group serves a control stimulus in this approach, the experiment was restricted to the remaining 16 groups. This design allows us to isolate responses to structural features beyond the shared power spectrum, including any symmetries other than translation, in the odd harmonics of the image update frequency \cite{RN1725,RN1540, RN1954}. Thus, the magnitude of the odd harmonic response components can be used as a distance metric for each group, with distance being measured relative to the simplest group, P1. A wallpaper group is a topologically discrete group of isometries of the Euclidean plane, i.e. transformations that preserve distance \cite{RN1425}. Wallpaper groups differ in the number and kind of these transformations. In mathematical group theory, when the elements of one group is completely contained in another, the inner group is called a subgroup of the outer group \cite{RN1425}. Subgroup relationships between wallpaper groups can be distinguished by their indices. The index of a subgroup relationship is the number of cosets, i.e. the number of times the subgroup is found in the outer group \cite{RN1425}. As an example, let us consider groups P6 and P2. If we ignore the translations in two directions that both groups share, group P6 consists of the set of rotations {$0^{\circ}$, $60^{\circ}$, $120^{\circ}$, $180^{\circ}$, $240^{\circ}$, $300^{\circ}$}, in which P2 {$0^{\circ}$, $180^{\circ}$} is contained. P2 is thus a subgroup of P6, and the full P6 set can be generated by every combination of P2 and rotations {$0^{\circ}$, $120^{\circ}$, $240^{\circ}$}. Because P2 is repeated three times in P6, P2 is a subgroup of P6 with index 3 \cite{RN1425}.

@@

>>>>

% If your first paragraph (i.e. with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation. If this is the case, add \parshape=0 to the end of the list environment.
\dropcap{S}ymmetries are present at many scales in images of natural scenes, due to a complex interplay of physical forces that govern pattern formation in nature. The importance of symmetry for visual perception has been known at least since the gestalt movement of the early 20th century. Since then, symmetry has been shown to contribute to the perception of shapes \cite{RN1311,RN1682}, scenes \cite{RN1824} and surface properties \cite{RN1166}, as well as the social process of mate selection \cite{RN1337}. Most of this work has focused on mirror symmetry or \textit{reflection}, with much less attention being paid to the other fundamental symmetries: \textit{rotation}, \textit{translation} and \textit{glide reflection}. In the two spatial dimensions relevant for images, these four fundamental symmetries can be combined in 17 distinct ways, the “wallpaper” groups \cite{RN1562,RN1563,RN1425}. Previous work has focused on four of the wallpaper groups, and used functional MRI to show that rotation symmetries within wallpapers are represented parametrically in several areas in occipital cortex, beginning with visual area V3 \cite{RN1725}. This effect is also robust in electroencephalography (EEG), whether measured using Steady-State Visual Evoked Potentials (SSVEPs)\cite{RN1725} or event-related paradigms \cite{RN1959}. Here we extend on this work by collecting SSVEPs and psychophysical data from human participants viewing the complete set of wallpaper groups. We measure responses in visual cortex to 16 out of the 17 wallpaper groups, with the 17th serving as a control stimulus, with the goal of providing a more complete picture of how wallpaper groups are represented in the human visual system.

A wallpaper group is a topologically discrete group of isometries of the Euclidean plane, i.e. transformations that preserve distance \cite{RN1425}. Wallpaper groups differ in the number and kind of these transformations. In mathematical group theory, when the elements of one group is completely contained in another, the inner group is called a subgroup of the outer group \cite{RN1425}. Subgroup relationships between wallpaper groups can be distinguished by their indices. The index of a subgroup relationship is the number of cosets, i.e. the number of times the subgroup is found in the outer group \cite{RN1425}. As an example, let us consider groups P6 and P2. If we ignore the translations in two directions that both groups share, group P6 consists of the set of rotations {$0^{\circ}$, $60^{\circ}$, $120^{\circ}$, $180^{\circ}$, $240^{\circ}$, $300^{\circ}$}, in which P2 {$0^{\circ}$, $180^{\circ}$} is contained. P2 is thus a subgroup of P6, and the full P6 set can be generated by every combination of P2 and rotations {$0^{\circ}$, $120^{\circ}$, $240^{\circ}$}. Because P2 is repeated three times in P6, P2 is a subgroup of P6 with index 3 \cite{RN1425}. The 17 wallpaper groups thus obey a hierarchy of complexity where simpler groups are sub-groups of more complex ones \cite{RN1711}.

The two datasets presented here puts on in the position of being able to assess the extent to which both behavior and brain responses follow that hierachy. The results show that activity in human visual cortex is remarkably consistent with the hierarchical relationships between the wallpaper groups, with SSVEP amplitudes and psychophysical thresholds preserving these relationships at a level that is far beyond chance. Visual cortex thus appears to encode all of the fundamental symmetries using a representational structure that closely approximates the subgroup relationships from group theory. Given that most participants had no knowledge of group theory, the ordered structure of visual responses to wallpaper groups is likely learned implicitly from regularities in the visual environment.

@@

The stimuli used in our two experiments were multiple exemplar images belonging to each of the wallpaper groups, generated from random-noise textures, as described in detail elsewhere \cite{RN1725}. Exemplar images from group P1 was used as control stimuli, and each exemplar from the other 16 groups had a power-spectrum matched P1 exemplar. The matched P1 exemplars were generated by phase-scrambling the exemplar images. Because all wallpapers are periodic due to their lattice tiling structure, the phase-scrambled images all belong to group P1 regardless of group membership of the original exemplar. P1 contains no symmetries other than translation, while all other groups contain translation in combination with one or more of the other three fundamental symmetries (reflection, rotation, glide reflection) \cite{RN1425}. In our SSVEP experiment, this stimulus set allowed us to isolate brain activity specific to the symmetry structure in the exemplar images from activity associated with modulation of low-level features, by alternating exemplar images and control exemplars. In this design, responses to structural features beyond the shared power spectrum, including any symmetries other than translation, are isolated in the odd harmonics of the image update frequency \cite{RN1725,RN1540, RN1954}. Thus, the combined magnitude of the odd harmonic response components can be used as a measure of the overall strength of the visual cortex response.

The psychophysical experiment took a distinct but related approach. In each trial an exemplar image was shown with its matched control, one image after the other, and the order varied pseudo-randomly such that in half the trials the original exemplar was shown first, and in the other half the control image was shown first. After each trial, participants were told to indicate whether the first or second image contained more structure, and the duration of both images was controlled by a staircase procedure so that a threshold duration could be computed for each wallpaper group.

A summary of our results can be seen in Figure 1 ...

@@

we used a steady-state design, in which exemplar images belonging to 16 of the 17 wallpaper groups alternated with phase-scrambled images of the same group. 
Exemplars from each of the 16 groups alternated at 0.83 Hz with their corresponding set of P1 exemplars, that were matched in terms of their Fourier power spectrum.

Thus, the magnitude of the odd harmonic response components can be used as a distance metric for each group, with distance being measured relative to the simplest group, P1.
